Donor Responses to Civil Society Repression

### 2025-01-23 index.qmd
Include hidden introduction heading for the TOC
Add revisions

---
title: "Donor Responses to Civil Society Repression"
subtitle: "Chapter for the *Handbook of Political Control*"
date: 2024-11-04
author:
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
  attributes:
    corresponding: true
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu

@@

::: {.content-visible when-format="wordcount-html"}
# Introduction {.hide}
:::

@@

The experiences of non-governmental organizations (NGOs) like AMERA are not uncommon. While a large literature in law and political science focuses on the role of international law, domestic legal institutions, and legal mobilization that may reduce state crackdown [@Klug:2005; @PowellStaton:2009; @Simmons:2009; @ConradRitter:2013], the last two decades have shown that states also use laws not to expand rights, but to further restrict rights and repress organizations [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @BakkeMitchellSmidt:2020; @GlasiusSchalkDeLange:2020; @BromleySchoferLonghofer:2020; @Chaudhry:2022]. Administrative crackdown—or the use of laws to create barriers to entry, funding, and advocacy by civil society groups—has proliferated across the globe over the past few decades—both across democracies and autocracies [@Chaudhry:2022]. As of 2023, 87% of the world's population now lives in countries where the space available to civil society is closed, repressed, or obstructed [@CIVICUSMonitor].

The growth in states using law to repress can be attributed to three main factors. First, administrative or legal crackdowns have many advantages over the use of violence. It allows governments to maintain a democratic facade and limit the domestic backlash because citizens tend to view legal crackdown as regulation rather than repression [@Chaudhry:2022]. Anti-NGO laws rarely elicit international condemnation or threats to withdraw aid. States also learn about the efficacy of these legal strategies from their peers and subsequently implement them in their own country [@GlasiusSchalkDeLange:2020; @Gilbert:2020; @Chaudhry:2022]. Finally, these tactics are part of a broader trend of using legal means to subvert democratic institutions and norms [@LevitskyZiblatt:2018].

@@

A rich literature on the purposes of foreign aid explores why states transfer money and resources to other countries. Both national aid agencies and multilateral international organizations face competing interests when deciding how to allocate resources, which projects to undertake, and which countries to work with. States rhetorically declare normative justifications for engaging in development, with recipient-focused goals like poverty reduction, democratization, human rights promotion, and environmental protection [@HoefflerOutram:2011]. In practice, though, aid is given for far more instrumental reasons based on donor countries' domestic and foreign policy-oriented strategic goals [@AlesinaDollar:2000; @Apodaca:2017]. A key enforcement mechanism to ensure that recipient-country actions align with donor-country goals is to condition aid on the achievement of specific democratizing and economic goals [@CheesemanSwedlundOBrien-Udry:2024; @HoefflerJustino:2024]. However, this dual-purpose strategy often creates dilemmas—if aid is withdrawn, it may jeopardize humanitarian programs or worsen political conditions. Conversely, continuing aid might signal tolerance for rights violations, weakening the credibility of the donor's human rights agenda. Prior research finds conflicting relationships between human rights abuses and foreign aid, variously concluding that donor agencies punish repressive states by reducing or withdrawing aid [@LebovicVoeten:2009], increasing aid to repressive states as a way to counter state violence [@Nielsen:2013], or making no measurable change in aid allocations [@Neumayer:2003; @Neumayer:2003b].

One reason for such mixed results is that the decision to withdraw or suspend aid in response to human rights abuses (including civil society crackdown) is often seen as a binary choice driven by self-interested motivations: a donor country that sees human rights abuses in a recipient state will consider its own foreign policy interests before deciding to either continue or curtail aid. In reality, however, the decision is far more complex and involves a wide range of possible motivations and responses. @CheesemanSwedlundOBrien-Udry:2024 offer a stylized decision tree of possible choices following recipient-country human rights violations—we provide a simplified version of their framework in @fig-agency-response-tree.

@@

Research on aid conditionality and agency responses to human rights abuses tends to look at more overt and violent forms of political repression. The effects of more subtle administrative crackdown on aid, including the passage of anti-NGO laws and the constricting of civic space, are less explored. Civil society plays a central role in the distribution and implementation of international aid. Government agencies, such as the United States Agency for International Development (USAID) and the United Kingdom's Foreign, Commonwealth and Development Office (FCDO), channel substantial funding through NGOs in recipient countries. Outsourcing the provision of aid to NGOs allows donors to avoid aid capture in weak institutional environments and can lead to more effective program implementation [@Dietrich:2013; @CruzLabonneWright:2024]. Given the importance of NGOs in aid delivery, we have reason to expect that donor responses to civil society crackdown abroad are treated differently than more violent forms of repression, following different pathways in @fig-agency-response-tree.

@@

One key reason for the divergence in the responses of national and international agencies is the degree to which these donors rely on recipient-country NGOs to implement programs. Donor agencies prioritize the openness of civil society in recipient countries in different ways. Multilateral donors like the World Bank and OECD tend to channel money directly to recipient governments rather than domestic NGOs and are thus more insulated from the effects of local civil society repression, which likely explains the lack of formal response to anti-NGO laws [@ChristensenWeinstein:2013; @DupuyPrakash:2018]. There is more variation, however, in how bilateral donors work with domestic NGOs. @AllenFlynn:2018 find that the domestic political environments of donor countries shape how those countries channel their aid abroad. Because they tend to focus more on poverty alleviation initiatives, states with left-leaning governments typically channel more aid through recipient-country NGOs. In contrast, right-leaning governments are typically interested in promoting the economic and political interests of their own states, and accordingly prefer to channel aid directly through recipient-country governments [@AllenFlynn:2018]. Human rights abuses in recipient countries will shape the responses of donor countries depending on the saliency of the threat to aid provision: political repression that threatens business interests and investments will be more salient to right-leaning governments, while anti-NGO civil society administrative crackdown will likely have a stronger effect on left-leaning governments' responses, since domestic NGOs are central to donor countries' aid distribution strategies.

@@

@@

Experimental evidence examines how donors react to the knowledge that potential recipient NGOs face legal difficulties. @ChaudhryHeiss:2021 find that foreign civil society restrictions serve as a heuristic when deciding to donate—learning about NGO repression increases individual generosity and causes potential donors to be willing to give substantially more to legally restricted nonprofits. This effect is strongest for privately-funded human rights-focused NGOs that face legal crackdown, as individuals feel that the organization deserves their support. Several participants justified their increased support because restricted NGOs are "doing good work in countries where it is tough for groups like them to operate and they need all the help they can get" [@ChaudhryHeiss:2021, 496]. Building on this work, @ChaudhryDotsonHeiss:2024 examine how organizational characteristics (e.g., issue area, funding sources) and practices (e.g., financial transparency) interact with structural concerns like legal crackdown abroad. They find that, all else equal, legal crackdown on its own causes donors to reduce their support for nonprofits—individuals are 5–10 percentage points less likely to donate to an NGO that is criticized or restricted, compared to an organization that enjoys a friendly relationship with its host government. However, this reduction in support can be offset by organizational characteristics. Financial transparency and accountability protect each increase the probability of donations by nine percentage points under the worst conditions of legal crackdown. Individual donors thus typically reduce their support when seeing legal crackdown, but change their response and increase their support when seeing other signs that the organization follows best practices and is more deserving.

@@

While private donations have become an important source of funding in the age of closing civic space and increased restrictions on foreign funding, they are unlikely to fully supplant large bilateral and multilateral sources of foreign aid, nor can they serve national foreign policy interests like more standard foreign aid. Testimony at a 2017 U.S. Senate Committee on Foreign Relations hearing maintained that "[n]either remittances nor philanthropy can replace the expertise, scale, or agenda setting capacity of American foreign policy assistance" [@Runde:2017]. Thus, despite promising new research on how to raise funds from private sources for besieged NGOs abroad, official aid donors also need to mount a coherent response to the shrinking civic space.

@@

While the examples of Kenya and Kyrgyzstan discussed above show that donors and local activists can successfully push back against anti-NGO laws, a more systematic approach is needed. To tackle this administrative repression, Western states need a more effective and coordinated response that engages not only aid agencies, donors, and international NGOs but also businesses, development communities, and local populations to address larger structural issues that exacerbate government crackdowns on civil society. Without such an effort, the repression of civil society organizations may be the canary in the coal mine that sets the groundwork for future democratic erosion.

>>>>

---
title: "Donor Responses to Civil Society Repression"
subtitle: "Chapter for the *Handbook of Political Control*"
date: 2025-01-23
author:
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
  attributes:
    corresponding: true
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu

@@

::: {.content-visible when-format="html"}
# Introduction {.hide}
:::

@@

The experiences of non-governmental organizations (NGOs) like AMERA are not uncommon. While a large literature in law and political science focuses on the role of international law, domestic legal institutions, and legal mobilization that may reduce state crackdown [@Klug:2005; @PowellStaton:2009; @Simmons:2009; @ConradRitter:2013], the last two decades have shown that states also use laws not to expand rights, but to further restrict rights and repress organizations [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @BakkeMitchellSmidt:2020; @GlasiusSchalkDeLange:2020; @BromleySchoferLonghofer:2020; @Chaudhry:2022]. Administrative crackdown—or the use of laws to create barriers to entry, funding, and advocacy by civil society groups—has proliferated across the globe over the past few decades—both across democracies and autocracies [@Chaudhry:2022]. As of 2023, 87% of the world's population now lives in countries where the space available to civil society is closed, repressed, or obstructed [@CIVICUSMonitor]. This intersection of administrative and legal institutions (layer 3) being used to control the activities and existence of civil society organizations (layer 4) is an example of how Earl and Braithwaite's [-@EarlBraithwaite:2022] layered model helps us consider the tools states use to control dialogue and mobilization among their populations.

The growth in states using law to repress civil society organizations can be attributed to three main factors. First, administrative or legal crackdown has many advantages over the use of violence. It allows governments to maintain a democratic facade and limit the domestic backlash because citizens tend to view legal crackdown as regulation rather than repression [@Chaudhry:2022]. Anti-NGO laws rarely elicit international condemnation or threats to withdraw aid. States also learn about the efficacy of these legal strategies from their peers and subsequently implement them in their own country [@GlasiusSchalkDeLange:2020; @Gilbert:2020; @Chaudhry:2022]. Finally, these tactics are part of a broader trend of using legal means to subvert democratic institutions and norms [@LevitskyZiblatt:2018].

@@

A rich literature on the purposes of foreign aid explores why states transfer money and resources to other countries. Both national aid agencies and multilateral international organizations face competing interests when deciding how to allocate resources, which projects to undertake, and which countries to work with. States rhetorically declare normative justifications for engaging in development, with recipient-focused goals like poverty reduction, democratization, human rights promotion, and environmental protection [@HoefflerOutram:2011]. In practice, though, aid is given for far more instrumental reasons based on donor countries' domestic and foreign policy-oriented strategic goals [@AlesinaDollar:2000; @Apodaca:2017]. A key enforcement mechanism to ensure that recipient-country actions align with donor-country goals is to condition aid on the achievement of specific democratizing and economic goals [@CheesemanSwedlundOBrien-Udry:2024; @HoefflerJustino:2024]. However, this dual-purpose strategy often creates dilemmas—if aid is withdrawn, it may jeopardize humanitarian programs or worsen political conditions. Conversely, continuing aid might signal tolerance for rights violations, weakening the credibility of the donor's human rights agenda. Prior research finds conflicting relationships between human rights abuses and foreign aid, variously concluding that donor agencies punish repressive states by reducing or withdrawing aid [@LebovicVoeten:2009], increasing aid to repressive states to counter state violence [@Nielsen:2013], or making no measurable change in aid allocations [@Neumayer:2003; @Neumayer:2003b].

One reason for such mixed results is that the decision to withdraw or suspend aid in response to human rights abuses (including civil society crackdown) is often seen as a binary choice driven by self-interested motivations: a donor country that sees human rights abuses in a recipient state will consider its own foreign policy interests before deciding to either continue or curtail aid. However, the decision is far more complex and involves a wide range of possible motivations and responses. @CheesemanSwedlundOBrien-Udry:2024 offer a stylized decision tree of possible choices following recipient-country human rights violations—we provide a simplified version of their framework in @fig-agency-response-tree.

@@

Research on aid conditionality and agency responses to human rights abuses tends to look at more overt and violent forms of political repression, or higher layers in the @EarlBraithwaite:2022 framework. The effects of more subtle administrative crackdown on aid—including the passage of anti-NGO laws, the constricting of civic space, and other layer 3 institutions—are less explored. Civil society plays a central role in the distribution and implementation of international aid. Government agencies, such as the United States Agency for International Development (USAID) and the United Kingdom's Foreign, Commonwealth and Development Office (FCDO), channel substantial funding through NGOs in recipient countries. Outsourcing the provision of aid to NGOs allows donors to avoid aid capture in weak institutional environments and can lead to more effective program implementation [@Dietrich:2013; @CruzLabonneWright:2024]. Given the importance of NGOs in aid delivery, we have reason to expect that donor responses to civil society crackdown abroad are treated differently than more violent forms of repression, following different pathways in @fig-agency-response-tree.

@@

One key reason for the divergence in the responses of national and international agencies is the degree to which these donors rely on recipient-country NGOs to implement programs. Donor agencies prioritize the openness of civil society in recipient countries in different ways. Multilateral donors like the World Bank and OECD tend to channel money directly to recipient governments rather than domestic NGOs and are thus more insulated from the effects of local civil society repression, which likely explains the lack of formal response to anti-NGO laws [@ChristensenWeinstein:2013; @DupuyPrakash:2018]. There is more variation, however, in how bilateral donors work with domestic NGOs. @AllenFlynn:2018 find that the domestic political environments of donor countries shape how those countries channel their aid abroad. Because they tend to focus more on poverty alleviation initiatives, states with left-leaning governments typically channel more aid through recipient-country NGOs. In contrast, right-leaning governments are typically interested in promoting the economic and political interests of their own states, and accordingly prefer to channel aid directly through recipient-country governments [@AllenFlynn:2018]. Human rights abuses in recipient countries will shape the responses of donor countries depending on the saliency of the threat to aid provision: political repression that threatens business interests and investments will be more salient to right-leaning governments, while anti-NGO civil society administrative crackdown will likely have a stronger effect on left-leaning governments' responses, since domestic NGOs are central to donor countries' aid distribution strategies. This dynamic indicates ways that the @EarlBraithwaite:2022 framework might be expanded to work between states—layer 3 institutions like civil society regulations in donor countries can affect perceptions of layer 3 institutions in recipient countries, and vice versa.

@@

This watering down of NGO programming illustrates interesting interplay between the different layers of Earl and Braithwaite's [-@EarlBraithwaite:2022] model of repression. Both de jure anti-NGO laws and the de facto civil society regulatory environment represent layer 3 institutions designed to limit participation in institutional politics. These limits cause civil society actors at layer 4 to shift to safer strategies and curtail their programming, which in turn can lead to worse consequences downstream. Formal and informal limits on civil society could be one mechanism behind the fact that foreign democracy aid increases the probability of civil war onset in states with weaker and less institutionalized civil society [@BraithwaiteLicht:2020]. Amid violent level 7 conflict, civil society restrictions may prevent NGOs from intervening effectively to address humanitarian challenges arising from violent conflict. For instance, Executive Order 13224—issued by George W. Bush in 2001—banned all financial transactions with any individual or organization on a list of Specially Designated Global Terrorists [@eo13224journal]. However, many humanitarian aid organizations (layer 4 institutions), work in environments where even the smallest circumstantial evidence could be grounds for designating these NGOs as having provided material support or resources for terrorism [@Bhungalia:2023]. Thus, many nonprofits may choose to avoid working in conflict zones. For example, after the US listed Al Shabaab in Somalia as a terrorist group, it resulted in a 88% reduction in civil society aid to Somalia over the next two years, with devastating consequences for Somalis [@Hayes:2017].

Other layers in this model of repression also point toward future research on donor responses to NGO restrictions. @BraithwaiteLicht:2020 find that states that receive foreign democracy assistance have a lower probability of internal conflict when civil society is more institutionalized. It is likely that broader social institutionalization (e.g., strong formal social institutions [level 5] and robust protest mobilization [level 6]) protects against more violent repression. We do not yet know how these other levels of institutionalization interact with level 3 civil society restriction, nor how official aid donors might condition their responses to repression. For instance, are donor states less likely to tame their programming when there are popular protests or powerful labor unions that can supplement resistance to government action?

@@

Experimental evidence examines how donors react to the knowledge that potential recipient NGOs face legal difficulties. @ChaudhryHeiss:2021 find that foreign civil society restrictions serve as a heuristic when deciding to donate—learning about NGO repression increases individual generosity and causes potential donors to be willing to give substantially more to legally restricted nonprofits. This effect is strongest for privately-funded human rights-focused NGOs that face legal crackdown, as individuals feel that the organization deserves their support. Several participants justified their increased support because restricted NGOs are "doing good work in countries where it is tough for groups like them to operate and they need all the help they can get" [@ChaudhryHeiss:2021, 496]. Building on this work, @ChaudhryDotsonHeiss:2024 examine how organizational characteristics (e.g., issue area, funding sources) and practices (e.g., financial transparency) interact with structural concerns like legal crackdown abroad. They find that, all else equal, legal crackdown on its own causes donors to reduce their support for nonprofits—individuals are less likely to donate to an NGO that is criticized or restricted, compared to an organization that enjoys a friendly relationship with its host government. However, this reduction in support can be offset by organizational characteristics. Financial transparency and accountability each increase the probability of donations under the worst conditions of legal crackdown. Individual donors thus typically reduce their support when seeing legal crackdown, but change their response and increase their support when seeing other signs that the organization follows best practices and is more deserving.

@@

While private donations have become an important source of funding in the age of closing civic space and increased restrictions on foreign funding, they are unlikely to fully supplant large bilateral and multilateral sources of foreign aid, nor can they serve national foreign policy interests like more standard foreign aid. Testimony at a 2017 U.S. Senate Committee on Foreign Relations hearing maintained that "[n]either remittances nor philanthropy can replace the expertise, scale, or agenda setting capacity of American foreign policy assistance" [@Runde:2017]. Thus, despite promising new research on how to raise funds from private sources for besieged NGOs abroad, official aid donors also need to mount a coherent response to the shrinking civic space, in which the actors comprising layer 4 of the @EarlBraithwaite:2022 model are challenged by restrictions from layer 3 institutions.

@@

While the examples of Kenya and Kyrgyzstan discussed above show that donors and local activists can successfully push back against anti-NGO laws, a more systematic approach is needed. To tackle this administrative repression, Western states need a more effective and coordinated response that engages not only aid agencies, donors, and international NGOs but also businesses, development communities, and local populations to address larger structural issues that exacerbate government crackdowns on civil society. This chapter demonstrates that the use of layer 3 controls---such as law and administrative repression---does not just have an impact on layer 4 actors such as civil society institutions. They also have far-reaching repercussions on efforts to address and resolve violent conflict (layer 7) and support for vulnerable communities, including minorities, queer groups, women, refugees, and asylum seekers (layer 2), who often depend on specialized nonprofit groups for services and legal advice that may cease to exist in the face of legal restrictions and anti-NGO rhetoric. Without such an effort, the repression of civil society organizations may be the canary in the coal mine that sets the groundwork for future democratic erosion.
